Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-09-10 paper/paper.tex
add manifest
update benchmarks

\section{Introduction}
\blue{overlap gap property~\cite{Gamarnik2013, Gamarnik2019}, independence polynomial at -1~\cite{Bousquet2008,Levit2009},
branching algorithm~\cite{Tarjan1977,Robson1986}
}
In graph theory and combinatorics, there are many interesting but hard computational problems concerning properties of independent sets.
An independent set is a set of vertices in a graph where no two vertices are adjacent to each other.
More formally, given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding an independent set of the largest possible size, the maximum independent set (MIS), is a paradigmatic NP-complete problem~\cite{Hastad1996}.
The size of the MIS of a graph $G$ is called the independence number, denoted as $\alpha(G) \equiv \max_{I}|I|$.

@@

This leads us to think about what is the most general element type that allowed in a tensor network contraction program.
Our finding is: as long as the tensor elements form a commutative semiring, the tensor network contraction result can be contraction order independent.
So what is a commutative semiring?
A semiring is a ring without additive inverse, while a commutative semiring is a semiring where the multiplication operation is commutative.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
The requirement of being commutative is for the tensor contraction result to be independent of the contraction order, which could be relaxed in certain cases. In the following sections, we show how to compute the independence polynomial,
the maximal independence polynomial, independence number, the number of independent sets, and enumeration of the MIS configurations of a general graph $G$ by designing tensor element types as commutative semirings while keeping the tensor network generic~\cite{Stepanov2014}.
Table~\ref{tbl:generictypes} summarizes the problems that can be solved by various tensor element types.

@@

\section{Benchmarks and case studies}
\subsection{Performance benchmarks}
We run a sequential program benchmark on CPU Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz, and show the results in Figure~\ref{fig:benchmark}.
Tensor network contraction can be parallelized. When the element type is immutable, one can also run it on GPUs to speed up the computation.
The graph we use are random 3 regular graphs, a typical type of sparse graph that having small tree width $tw(G) \leq |V|/6$~\cite{Fomin2006}.

\begin{figure} 
    \centering
    \includegraphics[width=0.8\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{benchmark.pdf}
    \caption{\red{label x-axis as ``number of vertices, $|V|$''} Benchmark results for computing different properties of independent sets with different tensor element types. The right axis is only for the dashed line.
    }\red{split into two graphs: one with time versus number of vertices, $|V|$, and one with time versus tree width.}
    \red{We also need to show/say what types of graphs is used here for benchmarking.}
    \label{fig:benchmark}
\end{figure}

>>>>

\section{Introduction}
\blue{overlap gap property~\cite{Gamarnik2013, Gamarnik2019}, independence polynomial at -1~\cite{Bousquet2008,Levit2009},
branching algorithm~\cite{Tarjan1977,Robson1986},
two motivations of computing independence polynomial: 1. Lee-Yang zero~\cite{Lee1952,Yang1952}, 2. Lov{\'a}sz local lemma~\cite{Scott2005}.
}
In graph theory and combinatorics, there are many interesting but hard computational problems concerning properties of independent sets.
An independent set is a set of vertices in a graph where no two vertices are adjacent to each other.
More formally, given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding an independent set of the largest possible size, the maximum independent set (MIS), is a paradigmatic NP-complete problem~\cite{Hastad1996}.
The size of the MIS of a graph $G$ is called the independence number, denoted as $\alpha(G) \equiv \max_{I}|I|$.

@@

This leads us to think about what is the most general element type that allowed in a tensor network contraction program.
Our finding is: as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result is contraction order independent.
A commutative semiring is a semiring with its multiplication operation being commutative, while a semiring is a ring without additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
The requirement of being commutative is for the tensor contraction result to be independent of the contraction order, which could be relaxed in certain cases. In the following sections, we show how to compute the independence polynomial,
the maximal independence polynomial, independence number, the number of independent sets, and enumeration of the MIS configurations of a general graph $G$ by designing tensor element types as commutative semirings while keeping the tensor network generic~\cite{Stepanov2014}.
Table~\ref{tbl:generictypes} summarizes the problems that can be solved by various tensor element types.

@@

\section{Benchmarks and case studies}
\subsection{Performance benchmarks}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100.
The results are shown in Figure~\ref{fig:benchmark}.
The graph we use are random 3 regular graphs, a typical type of sparse graph that having small tree width $tw(G) \leq |V|/6$~\cite{Fomin2006}.

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig1.pdf}
    \caption{Benchmark results for computing different properties of independent sets on a random three regular graph with different tensor element types.
    (b) is the computing time for computing maximum independent set size, number of independent sets and suboptimal sets.
    (c) is the computing time for independence polynomials with different approaches.
    (d) is the computing time for finding all configurations and a single configuration, with or without bounding.
    }\red{split into two graphs: one with time versus number of vertices, $|V|$, and one with time versus tree width.}
    \red{We also need to show/say what types of graphs is used here for benchmarking.}
    \label{fig:benchmark}
\end{figure}

Among the benchmarks shown in \Fig{fig:benchmark}, computing \texttt{max size (CPU)},
\texttt{total size (CPU)}, \texttt{total size (GPU)}, \texttt{IDP (FFT) (CPU)}, \texttt{IDP (FFT) (GPU)} and \texttt{single configuration (bounding) (CPU)} can utilize fast BLAS libraries,
hence they are relatively faster comparing to other methods in the same category.
Only algebras requiring dynamic sized structures can not be computed on GPU, here are polynomial algebra for computing independence polynomial and set algebra for enumerating configurations.
When computing the independence polynomial, the finite field approach is the only method without round off error that can run on GPU.
